
\chapter{Conclusion and Further Research}

\section{Derivative-Free versus Derivative-Based Methods}
We have shown that use of derivative free methods can allow for a considerable reduction in time spent on writing code to fit FDA problems. Brent's Method is much simpler than \texttt{Data2LD.opt}.

The derivative-free approch is effective at fitting data using non-smooth loss functions such as the sum of absolute deviations. A huge advangtage is that it allows one to piggyback on existing code that only computes a goodness-of-fit statistic for a given choice of parameters. The problem of numerically finding the function that minimises a Penalised Sum of Squares criterion is outside the mainstream of Numerical Analysis, which means that one does not have a body of work to fall back on. One should thus try to avoid having to write solvers for the inner problem if possible.

The derivative-free approach further allows problems that are not mainstream within FDA  such as parameterically fitting PDEs to be solved without having to implement a complex solver like \texttt{Data2LD.}

However, the methodology suffers from some weaknessess. It is undeniably slower than derivative-based methods. The MM algorithm for $L_1$ fitting suffers particularly badly from this weakness because it must solve many weighted $PENSSE$ problems in the course of its execution and it is difficult to accelerate. This must be balanced against the MM alogrithm's ease of implemention. 

\section{Further Subjects for Inquiry}

An obvious topic for further investigation is combining Brent's method with \texttt{Data2LD} so that the latter package can handle the question of parameter estimation, while Brent's method is used to find the optimal value of $\lambda.$

Another question to be asked  is whether the MM alogrithm for $L_1$ fitting can be combined with \texttt{Data2LD} to perform $L_1$ fitting where the parameters of the ODE are time varying functions.

Finally, while  the two level parameter cascade with $L_1$ errors was implemented, no effort was made to implement the three level parameter cascade. It has not been investigated by us whether the GCV is appropriate as a fitting criterion for $\lambda$ in the $L_1$ case. 


\section{Quasi-linear Differential Problems}

Throughout this thesis, it has been possible to either use techniques from Applied Mathematics to construct solution strategies on a case by case basis, or use Statistical methods to find a semi-parametric fit. As differential equations become more complex, both approaches begin to rapidly become non-viable. In this section,  quasi-linear differential equation models will be touched upon to illustrate how small changes can greatly increase the difficulty of fitting.

The difference between a quasi-linear and a linear differential equation is that the coefficients in a quasi-linear equation are allowed to depend on the unknown function\footnote{In practice there is no precise definition of `quasi-linear', it is defined as suits the problem domain at hand}. Instead of an ODE such as $y' = \beta(t)y,$ one would have an ODE such as $y' = \beta(y,t)y.$  Though quasi-linear problems tend to be reminiscent of linear ones, they are nonetheless substanially more complicated, and require more technical knowledge, and even ingenuity to tackle.

For a quasi-linear variation of a linear ODE, consider the Van Der Pol Equation:

\[
   y''(t) + \beta (1 - y(t))^2 y'(t) + y(t)
\]

This ODE has no obvious solution.


\subsection{Inviscid Burger's Equation}
Even if a solution exists, an estimation strategy might be difficult to derive. Consider the inviscid Burger's Equation:
\[
    \frac{\partial u(x,t)}{\partial t} + \beta u(x,t) \frac{\partial u(x,t)}{\partial x} = 0
\]

This equation is identical to  the Transport Equation except that the rate term is equal to $\beta u(x,t).$ The solution is given by:

\[
   u(x,t) = f(s)
\]

Here $f(\cdot)$ is some arbitary function as  before, and $s$ is implicitly defined as the solution of the equation $x = \beta f(s) t + s$. Since $s = x - \beta ut,$ this can be written as:

\[
   u(x,t) = f(x - \beta ut)
\]


Fitting this model is substantially trickier than the Transport Equation. There is no clean seperation between the problem of estimating $f(\cdot)$ and $\beta$ since $u(x,t)$ appears on the righthand side and scales $\beta.$ 

A further complication is that $u(x,t)$ might only define a \emph{relation}, instead of a function. There might be multiple values of $u$ associated with a given  $(x,t)$ that satisfy the solution equation. Physically speaking, multiple values correspond to shock waves. 

\subsection{Discussion} 

We see that the level of knowledge required to devise fitting strategies can increase substantially even with seemingly modest increases in the complexity of the associated differential equation.


Consider the following quasi-linear model of genetic drift in a population proposed\footnote{In the \textit{Annals of Eugenics...}} by R.A. Fisher:\cite{fisher1937wave}

\begin{equation}
\frac{\partial u(x,t)}{\partial t} + \beta_1 \frac{\partial u(x,t)}{\partial x} = \beta_2 u(x,t)(1 - u(x,t))
\end{equation}

This problem is similar to the previous  PDEs we discussed, it even admits travelling wave solutions of the form $f(x + ct)$ as Fisher himself noted. Nonetheless, it is a much more difficult problem despite the apparently modest increase in complexity. One would likely  have to consult a textbook that covers non-linear PDEs that can generate waves in fair degree of detail to be able to devise a fitting strategy. This is quite a specialised subject!

The overall result is that as the complexity of the differential equation increases, more and more time will be needed to model it correctly.


